---
title: "Homework 3 sc5558"
author: "Shiyun Angel Cheng"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 0 

```{r load_libraries}
library(tidyverse)
library(janitor)
library(lubridate)
library(gt)
library(patchwork)
library(scales)
```
## Problem 1 

This problem uses the Instacart data. DO NOT include this dataset in your local data directory; instead, load the data from the p8105.datasets using:
```{r}
library(p8105.datasets)
data("instacart")
```

The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illstrative examples of observations.

Now I am conducting an initial exploraion of the dataset 
```{r}
instacart = instacart |>
  janitor::clean_names()

instacart |> glimpse()
```
This dataset contains grocery orders from Instacart. Each observation represents a single product purchased by a user in a single order. The dataset has `r nrow(instacart)` observations and `r ncol(instacart)` variables. Key variables include:

product_name, aisle, and department: describe the purchased product;

order_hour_of_day and order_dow: describe when the order was placed;

add_to_cart_order and reordered: describe the user’s shopping behavior.

Now we move on to answering the following quesitons: 
1. How many aisles are there, and which aisles are the most items ordered from?
```{r}
aisle_counts = instacart |>
  count(aisle, name = "n_items") |>
  arrange(desc(n_items))

n_aisles = nrow(aisle_counts)
top_aisles = aisle_counts |> slice_max(n_items, n = 5)

n_aisles
```
There are `r n_aisles` aisles in total. The top 5 aisles by number of items ordered are below: 
```{r}
top_aisles |>
  gt() |>
  cols_label(
    aisle = "Aisle",
    n_items = "Items ordered"
  ) |>
  tab_header(title = "Top 5 Aisles by Items Ordered")
```
2. Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.
```{r}
instacart |>
  count(aisle, name = "n_items") |>
  filter(n_items > 10000) |>
  mutate(aisle = fct_reorder(aisle, n_items)) |>
  ggplot(aes(x = aisle, y = n_items)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Aisles with More Than 10,000 Items Ordered",
    x = NULL,
    y = "Items ordered"
  ) +
  theme_minimal()
```
This plot emphasizes that fresh produce and dairy products are ordered most. 

3. Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.
```{r}
selected_aisles = c("baking ingredients", "dog food care", "packaged vegetables fruits")

instacart |>
  filter(aisle %in% selected_aisles) |>
  count(aisle, product_name, name = "n") |>
  group_by(aisle) |>
  slice_max(n, n = 3) |>
  ungroup() |>
  arrange(aisle, desc(n)) |>
  gt(groupname_col = "aisle") |>
  cols_label(
    product_name = "Product",
    n = "Times ordered"
  ) |>
  tab_header(title = "Top 3 Products in Selected Aisles")
```
Packaged vegetables fruits aisle are ordered most times, with baking ingredients and dog food care following. For packaged vegetables fruits, organic baby spinach and raspberries are ordered most.For baking ingredients, brown sugar, baking soda, and cane sugar are ordered most. For dog food care, dog treats, recipe, and dog biscuits are ordered most. 

4. Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).
```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day), .groups = "drop") |>
  mutate(order_dow = factor(order_dow, levels = 0:6,
                            labels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))) |>
  pivot_wider(names_from = order_dow, values_from = mean_hour) |>
  gt(rowname_col = "product_name") |>
  fmt_number(everything(), decimals = 2) |>
  tab_header(title = "Mean Order Hour by Day of Week")
```
This table shows that Pink Lady Apples are generally ordered earlier in the day, while Coffee Ice Cream tends to be ordered slightly later. 

### Problem 2 

This Problem uses the Zillow datasets introduced in Homework 2. Both datasets are available here. Import, clean, and otherwise tidy these datasets.
```{r}
# Load and clean Zillow data (wide format)
zori_df = read_csv("./data/zilow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv") |>
  janitor::clean_names() |>
  pivot_longer(
    cols = matches("^20"),
    names_to = "ym",
    values_to = "zori"
  ) |>
  rename(zip = region_name) |>
  mutate(
    zip = as.character(zip),
    ym = lubridate::ym(ym),
    year = lubridate::year(ym)
  )

# Load and clean ZIP–borough lookup
zipcode_df = read_csv("./data/zilow_data/Zip Codes.csv") |>
  janitor::clean_names() |>
  rename(zip = zipcode) |>
  mutate(zip = as.character(zip))

# Merge ZORI data with ZIP code lookup
zori_df = zori_df |>
  inner_join(zipcode_df, by = "zip") |>
  relocate(county, neighborhood, .after = zip)
```


There are 116 months between January 2015 and August 2024. How many ZIP codes are observed 116 times? How many are observed fewer than 10 times? Why are some ZIP codes are observed rarely and others observed in each month?

Create a reader-friendly table showing the average rental price in each borough and year (not month). Comment on trends in this table.

Make a plot showing NYC Rental Prices within ZIP codes for all available years. Your plot should facilitate comparisons across boroughs. Comment on any significant elements of this plot.

Compute the average rental price within each ZIP code over each month in 2023. Make a reader-friendly plot showing the distribution of ZIP-code-level rental prices across boroughs; put differently, your plot should facilitate the comparison of the distribution of average rental prices across boroughs. Comment on this plot.

Combine the two previous plots into a single graphic, and export this to a results folder in your repository.
